---
title: "Heart Failure Prediction – ensemble models"
author: "Nancy Walker"
format: 
    html: 
        toc: true
    pdf: default
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document outlines and evaluates the ensemble modeling on the preprocessed Heart Failure Prediction data set. Ensemble models used include decision trees, k-nearest neighbors, and neural networks.

## 1. Load Required Libraries

```{r}
library(caret)
library(pROC)

set.seed(100)
```

## 2. Import Training and Testing Data sets

```{r}
heart_train <- read.csv("heart_preprocessed_training.csv")
heart_test <- read.csv("heart_preprocessed_test.csv")
```

## 4. Prepare data for Modeling 

```{r}
# Convert HeartDisease to factor 
heart_train$HeartDisease <- factor(heart_train$HeartDisease, levels = c("No", "Yes"))
heart_test$HeartDisease <- factor(heart_test$HeartDisease, levels = c("No", "Yes"))

# Convert predictors to factor 
#train
heart_train$Sex <- as.factor(heart_train$Sex)
heart_train$ChestPainType <- as.factor(heart_train$ChestPainType)
heart_train$RestingECG <- as.factor(heart_train$RestingECG)
heart_train$ExerciseAngina <- as.factor(heart_train$ExerciseAngina)
heart_train$ST_Slope <- as.factor(heart_train$ST_Slope)
#test 
heart_test$Sex <- as.factor(heart_test$Sex)
heart_test$ChestPainType <- as.factor(heart_test$ChestPainType)
heart_test$RestingECG <- as.factor(heart_test$RestingECG)
heart_test$ExerciseAngina <- as.factor(heart_test$ExerciseAngina)
heart_test$ST_Slope <- as.factor(heart_test$ST_Slope)

#Define predictors and independent variable 
x = heart_train[,1:11]
y = heart_train$HeartDisease

#Create Folds to enable resampling 
set.seed(100)
folds <- createFolds(y, returnTrain = TRUE)
#ensure consistant resampling
ctrl <- trainControl(method = "repeatedcv", number = 10,
                     repeats = 5,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = "final")
```

## 3. Ensemble Models

### Decision Tree

```{r}
#Decision Tree 
dtModel <- train(x = x, y= y,
                 method = "rpart",
                 tuneLength = 30,
                 metric = "ROC",
                 trControl = ctrl)

dtModel
ggplot(dtModel)
testResults <- data.frame(obs = heart_test$HeartDisease,
                          dt = predict(dtModel, heart_test[, 1:11]))
```

### K-Nearest Neighbors

```{r}
#KNN
#remove non-zero variance to reduce noise 
knnDescr <- x[, -nearZeroVar(x)] 
knnTestDescr <- heart_test[, colnames(knnDescr)]

set.seed(100)
knnModel <- train(x = knnDescr, y = y,
                 method = "knn",
                 tuneGrid = data.frame(k = 1:20),
                 metric = "ROC",
                 trControl = ctrl)
knnModel
ggplot(knnModel)
testResults$KNN <- predict(knnModel, knnTestDescr)
```

### Neural Network

```{r}
#Neural Networks 
#hyper parameter grid 
nnetGrid <- expand.grid(#regularize coefficents,
                        decay = c(0, 0.01, .1), 
                        #Unites in a hidden layer
                        size = c(1:10))

set.seed(100)
nnModel <- train(x, y,
                 method = "nnet",
                 tuneGrid = nnetGrid,
                 metric = "ROC",
                 trControl = ctrl,
                 trace = FALSE, maxit =500)
nnModel
ggplot(nnModel)
 
testResults$NN <- predict(nnModel, heart_test[, 1:11])
```

## 4. Test Results 

```{r}
train_metrics <- resamples(list(DT = dtModel,
    KNN = knnModel,
    NN = nnModel))

summary(train_metrics)

diff(train_metrics) |> summary()
```

## 5. ROC AUC 

```{r, fig.width=8, fig.height=7}
#Decision Tree
dtRoc <- roc(response = dtModel$pred$obs,
             predictor = dtModel$pred$No,
             levels = rev(levels(dtModel$pred$obs)))
#K-Nearest Neighbors
knnRoc <- roc(response = knnModel$pred$obs,
             predictor = knnModel$pred$No,
             levels = rev(levels(knnModel$pred$obs)))
#Neural Network
nnRoc <- roc(response = nnModel$pred$obs,
             predictor = nnModel$pred$No,
             levels = rev(levels(nnModel$pred$obs)))

### Compare Models using ROC curve
plot(dtRoc, type = "s", col = 'red', legacy.axes = TRUE)
plot(knnRoc, type = "s", add = TRUE, col = 'green', legacy.axes = TRUE)
plot(nnRoc, type = "s", add = TRUE, col = 'blue', legacy.axes = TRUE)
legend("bottomright", legend=c("DT", "KNN", "NN"),
       col=c("red", "green","blue"), lwd=2)
title(main = "Compare ROC curves from different models", outer = TRUE, 
      line = -1)
```

## Accuracy 

```{r}
accuracy <- function(true, pred) mean(true == pred)

cat("DT accuracy:", accuracy(testResults$obs, testResults$dt), "\n")
cat("KNN accuracy:", accuracy(testResults$obs, testResults$KNN), "\n")
cat("NN accuracy:", accuracy(testResults$obs, testResults$NN), "\n")

```

## Conclusion 

Neural networks had the largest ROC(AUC) with a median ROC(AUC) of 0.92, followed by decision trees that had a median ROC(AUC) of 0.89. K-nearest neighbors have the lowest median ROC(AUC) of 0.50.

Neural networks also had the greatest median sensitivity of 0.80, followed by decision trees with a median sensitivity of 0.78. K-nearest neighbors had a median sensitivity of 0.00. K-nearest neighbors had perfect specificity.

Neural networks had a median specificity of 0.90, followed by decision trees with a median specificity of 0.86.

Decision trees have the highest accuracy of 0.83, followed by neural networks with an accuracy of 0.63. K-nearest neighbors have the lowest accuracy of 0.55.

Decision trees and neural network metric are statistically significant using a 95% confidence interval when compared with k-nearest networks. Both have p-values of 2.2e-16 for ROC, sensitivity, and specificity. The difference between decision trees and neural networks is significant but small with p-values of 6.6e-06, 0.098, and 2.06e-05 for ROC, sensitivity, and specificity respectively.

Overall, neural networks have the best predictive ability because it has the highest ROC, balanced sensitivity & specificity.
