---
title: "Heart Failure Prediction – Data Preprocessing"
author: "Team 4: Jorge Roldan, Nancy Walker, Alli McKernan"
format: pdf
editor: visual
---

## Introduction

This document outlines the data cleaning and preprocessing steps performed on the Heart Failure Prediction dataset.

## 1. Load Required Libraries

We begin by loading necessary packages for data wrangling, preprocessing, and visualization.

```{r}
library(tidyverse)
library(caret)
library(DataExplorer)
```

## 2. Load and Inspect the Dataset

We load the heart.csv dataset and conduct an initial inspection of its structure and contents.

```{r}
heart_data <- read.csv("heart.csv")

# View structure and summary
str(heart_data)
summary(heart_data)
```

## 3. Check for Missing Values

It's important to verify that the dataset does not contain missing values that could interfere with modeling.

```{r}
sapply(heart_data, function(x) sum(is.na(x)))
```

## 4. Convert Categorical Variables

Categorical variables are converted to factors to ensure compatibility with modeling algorithms.

```{r}
heart_data$HeartDisease <- factor(heart_data$HeartDisease, levels = c(0, 1), labels = c("No", "Yes"))
heart_data$Sex <- as.factor(heart_data$Sex)
heart_data$ChestPainType <- as.factor(heart_data$ChestPainType)
heart_data$RestingECG <- as.factor(heart_data$RestingECG)
heart_data$ExerciseAngina <- as.factor(heart_data$ExerciseAngina)
heart_data$ST_Slope <- as.factor(heart_data$ST_Slope)
```

## 5. Explore for Low-Variance Features

Features with near-zero variance do not contribute much to prediction and may be dropped.

```{r}
nzv <- nearZeroVar(heart_data, saveMetrics = TRUE)
nzv
```

## 6. Split the Data into Training and Testing Data sets 

```{r}
#create training dataset for heart_data_scaled #80% of data to the trianing dataset 
set.seed(100) 
trainingRows <- createDataPartition(heart_data$HeartDisease, p = .80, list = FALSE)

#subset training and test datasets 
trainingData <- heart_data[trainingRows,] 
testingData <- heart_data[-trainingRows,]
```

## 6. Normalize Numeric Features

We scale and center the numeric variables to prepare them for algorithms sensitive to feature scale.

```{r}
numeric_features <- trainingData %>%
  select(where(is.numeric)) %>%
  names()

preproc <- preProcess(trainingData[, numeric_features], method = c("center", "scale"))
heart_data_train_scaled <- predict(preproc, heart_data)
```

## 7. Final Data Check

We perform a final check to ensure the cleaned dataset is ready for modeling.

```{r}
str(heart_data_train_scaled)
summary(heart_data_train_scaled)
```

## 8. Export the Cleaned Data set

```{r}
write.csv(heart_data_train_scaled, "heart_preprocessed_training.csv", row.names = FALSE)
write.csv(testingData,"heart_preprocessed_test.csv", row.names = FALSE )

```
